---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# groupdata2

R package: Subsetting Methods for Balanced Cross-Validation, 
Time Series Windowing, and General Grouping and Splitting of Data.  

By Ludvig R. Olsen  
Cognitive Science, Aarhus University  
Started in Oct. 2016 

Contact at:
r-pkgs@ludvigolsen.dk

Main functions:  
* group_factor  
* group  
* splt  
* fold  
  
Other tools:  
* %staircase%  

## Installation  
Development version:  
install.packages("devtools")  
devtools::install_github("LudvigOlsen/groupdata2")  

## To do  
* Refine vignettes 
* fold() - implement force_equal (n.b. should be special for greedy and staircasing)  
* datatables 
* Find suitable license
* Change version number
* Send to CRAN

## Features  
Write up a motivation for why you would want to use groupdata2.  

force_equal  
randomize  
etc.

## Examples

```{r error=FALSE,warning=FALSE,message=FALSE}
# Attach packages
library(groupdata2)
library(dplyr)
library(knitr)

```

### group()

```{r}
# Create dataframe
df <- data.frame("x"=c(1:12),
  "species" = rep(c('cat','pig', 'human'), 4),
  "age" = sample(c(1:100), 12))

# Using group()
group(df, 5, method = 'n_dist') %>%
  kable()

# Using group() with dplyr pipeline to get mean age
df %>%
  group(5, method = 'n_dist') %>%
  dplyr::summarise(mean_age = mean(age)) %>%
  kable()

```

### fold()

```{r}
# Create dataframe
df <- data.frame(
  "participant" = factor(rep(c('1','2', '3', '4', '5', '6'), 3)),
  "age" = rep(sample(c(1:100), 6), 3),
  "diagnosis" = rep(c('a', 'b', 'a', 'a', 'b', 'b'), 3),
  "score" = sample(c(1:100), 3*6))
df <- df[order(df$participant),]
df$session <- rep(c('1','2', '3'), 6)

# Using fold()

# With cat_col and id_col
df_folded <- fold(df, 3, cat_col = 'diagnosis',
                  id_col = 'participant', method = 'n_dist')

# Show df_folded ordered by folds
df_folded[order(df_folded$.folds),] %>%
  kable()

# Show distribution of diagnoses and participants
df_folded %>% 
  group_by(.folds) %>% 
  count(diagnosis, participant) %>% 
  kable()


```

