---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# groupdata2

R package: Subsetting Methods for Balanced Cross-Validation, 
Time Series Windowing, and General Grouping and Splitting of Data.  

By Ludvig R. Olsen  
Cognitive Science, Aarhus University  
Started in Oct. 2016 

Contact at:
r-pkgs@ludvigolsen.dk

Main functions:  
* group_factor  
* group  
* splt  
* fold  
  
Other tools:  
* %staircase%  

## Installation  
Development version:  
install.packages("devtools")  
devtools::install_github("LudvigOlsen/groupdata2")  

## To do  
* fold() - implement force_equal (n.b. should be special for greedy and staircasing)  
* datatables 
* Change version number
* Send to CRAN

## Functions  
### group_factor()  
Returns a factor with group numbers, e.g. 111222333.  
This can be used to subset, aggregate, group_by, etc.   

Create equally sized groups by setting force_equal = TRUE  
Randomize grouping factor by setting randomize = TRUE  

### group()  
Returns the given data as a dataframe with added grouping factor made with group_factor(). The dataframe is grouped by the grouping factor for easy use with dplyr pipelines.  

### splt()  
Splits the given data into the specified groups made with group_factor() and returns them in a list.  

### fold()  
Creates (optionally) balanced folds for use in cross-validation. Balance folds on one categorical variable and/or make sure that all datapoints sharing an ID is in the same fold.  
 
## Methods
There are currently 6 methods available. They can be divided into 3 categories.  
Examples of group sizes are based on a vector with 57 elements.  

### Specify group size
##### Method: greedy
Divides up the data greedily given a specified group size.  
E.g. group sizes: 10, 10, 10, 10, 10, 7   

### Specify number of groups
##### Method: n_dist (Default)
Divides the data into a specified number of groups and 
distributes excess data points across groups.  
E.g. group sizes: 11, 11, 12, 11, 12  

##### Method: n_fill
Divides the data into a specified number of groups and 
fills up groups with excess data points from the beginning.  
E.g. group sizes: 12, 12, 11, 11, 11  

##### Method: n_last
Divides the data into a specified number of groups. 
The algorithm finds the most equal group sizes possible, 
using all data points. Only the last group is able to differ in size.  
E.g. group sizes: 11, 11, 11, 11, 13  

##### Method: n_rand
Divides the data into a specified number of groups. 
Excess data points are placed randomly in groups (only 1 per group).  
E.g. group sizes: 12, 11, 11, 11, 12  

### Specify step size
##### Method: staircase
Uses step_size to divide up the data. 
Group size increases with 1 step for every group, until there is no more data.  
E.g. group sizes: 5, 10, 15, 20, 7


## Examples

```{r error=FALSE,warning=FALSE,message=FALSE}
# Attach packages
library(groupdata2)
library(dplyr)
library(knitr)

```

```{r}
# Create dataframe
df <- data.frame("x"=c(1:12),
  "species" = rep(c('cat','pig', 'human'), 4),
  "age" = sample(c(1:100), 12))

```


### group()

```{r}
# Using group()
group(df, 5, method = 'n_dist') %>%
  kable()

# Using group() with dplyr pipeline to get mean age
df %>%
  group(5, method = 'n_dist') %>%
  dplyr::summarise(mean_age = mean(age)) %>%
  kable()

```

### fold()

```{r}
# Create dataframe
df <- data.frame(
  "participant" = factor(rep(c('1','2', '3', '4', '5', '6'), 3)),
  "age" = rep(c(20,23,27,21,32,31), 3),
  "diagnosis" = rep(c('a', 'b', 'a', 'b', 'b', 'a'), 3),
  "score" = c(10,24,15,35,24,14,24,40,30,50,54,25,45,67,40,78,62,30))
df <- df[order(df$participant),]
df$session <- rep(c('1','2', '3'), 6)

```

```{r}
# Using fold() 

# First set seed to ensure reproducibility
set.seed(1)

# Use fold() with cat_col and id_col
df_folded <- fold(df, 3, cat_col = 'diagnosis',
                  id_col = 'participant', method = 'n_dist')

# Show df_folded ordered by folds
df_folded[order(df_folded$.folds),] %>%
  kable()

# Show distribution of diagnoses and participants
df_folded %>% 
  group_by(.folds) %>% 
  count(diagnosis, participant) %>% 
  kable()

```

